Robust inference of neutron-star parameters from thermonuclear burst

### 2021-09-17 doc/ms.tex
Lots of small changes to improve robustness, consistency
Introduced intvl_to_errors to convert the percentiles output from functions to proper 2-sided errors as used in input.
Corrected the way value_to_dist deals with scalars, as well as how it treats units with or without prefactors.
Changed the warning, error messages to use the logger instead
Updates to several functions to improve consistency of behaviour

Thermonuclear (type-I) bursts are observed from neutron stars accreting from low-mass ($\lesssim 1\ M_\odot$) binary companions \cite[e.g.][]{gal17b}. ...

@@

[{\bf insert table with key burst parameters here? }]

@@

One of the most significant astrophysical uncertainties that has largely been neglected to date is the anisotropy of the burst (and persistent) emission.
% [{\bf move discussion from later section to here?}]
% \subsection{Emission anisotropy}
% \label{subsec:anisotropy}
[{\bf bit more description here }]

@@

From data taken prior to the burst ignition we can measure the persistent flux $F_{\rm per}$, a measure of the accretion rate; [{\bf add expression here}]
% and provided more than one burst is observed, we may also measure (or estimate) the recurrence time, $\Delta t$. 
the persistent spectra of burst sources are  more complex than the bursts themselves, and typically exhibit thermal as well as non-thermal components [{\bf ref}].
%
These parameters may be combined to estimate the $\alpha$-value, the (observed) ratio of the burst to persistent flux: 
\begin{equation}
\alpha = \frac{\Delta t F_{\rm per} c_{\rm bol}}{E_B}
\end{equation}
where $c_{\rm bol}$ is the bolometric correction giving the inverse fraction of total persistent flux emitted in the instrumental band.
%
The $\alpha$ parameter is an important diagnostic of the burst fuel and ignition conditions.
%
[{\bf mention the average method, as used in the rates paper }]

@@

Of the model input parameters, we can also infer the persistent flux level expected given the model-assumed accretion rate $\dot{m}$, as follows: % gal03d, equation 2
\begin{equation}
% function fper in burstclass.py
F_{p,\infty} = \frac{L_p}{4\pi d^2} = 
% \frac{\dot{M} Q_{\rm grav}}{1+z} \left(\frac{R_{\rm NS}}{d}\right)^2 \xi_p^{-1}
% expression used in code
\frac{\dot{m} Q_{\rm grav}}{4\pi d^2 (1+z)\xi_p c_{\rm bol}}
\end{equation}
% Q_grav already defined in the previous section
% where $Q_{\rm grav} = c^2z/(1+z) \approx GM_{\rm NS}/R_{\rm NS}$ is the gravitational energy release per gram. % gal03d 
%
The bolometric correction $c_{\rm bol}$ accounts for the expermental limitation that the persistent flux can only be measured over a limited instrumental passband. The bolometric correction is the ratio of the estimated bolometric flux to the band-limited value.

@@

Describe how the code is written and how it works, the astropy.distribution thingy, MCMC for error determination etc.

@@

Here we describe different approaches that have been used to deduce system parameters from observations of thermonuclear bursts.
%
We focus on adopting a suitable approach given the availability of data, which will differ for different sources. 
%
We also describe how the {\sc concord} code can be used to fully account for the astrophysical uncertainties, and as a companion to this section (and the code itself) we provide a companion {\sc jupyter} notebook [{\bf ref}] which replicates the analyses described below.

@@

The distance can be estimated from the peak flux of a PRE burst based on equation \ref{disteq}. The challenge is taking into account the uncertainties in the various parameters that go into this calculation. If we adopt the empirical Eddington luminosity of \cite[]{kuul03a}, the only remaining factor is the burst emission anisotropy $\xi_b$.
%
As for the case of the burst in IGR~J00291+5934, the approach adopted in {\sc concord} to take this into account is to generate a distribution of $\xi_b$ values, based on an isotropic distribution of inclinations within known limits. The presence (absence) of X-ray dips occurring at the binary orbital period will constrain the inclination to greater (less) than $75^\circ$ [{\bf ref}]. However, more stringent constraints may be available.

@@

\cite{kuiper20} estimated the distance to the source based on the empirical value of the Eddington luminosity, and incorporating the expected anisotropy for an inclination range of 
% kuiper20 uses slightly more restrictive 28-30
24--30$^\circ$ \cite[]{sanna18} as $7.7_{-0.6}^{+0.8}$~kpc. Here the uncertainty is dominated by the peak flux, since the anisotropy factor is effectively fixed for such a narrow range of inclination, at $\xi_b=0.70$.
%
The PDF for the distance, which is distinctly non-Gaussian, is plotted in Figure \ref{fig:dist_igr17591}. We also show the corresponding distribution resulting from adoption of the theoretical Eddington luminosity, according to equation \ref{ledd}.

@@

The {\sc concord} code allows us to pass the derived distance distribution directly to the method for calculating the accretion rate at the time of the burst, also incorporating the dependence of the persistent flux anisotropy (which is not the same as for the burst flux).
%
The persistent flux can then be converted to an accretion rate, given the estimated anisotropy factor predicted (separately) for the persistent emission, as
\begin{eqnarray}
\dot{m} & = L_{\rm pers} (1+z) (4\pi R^2(GM/R_{\rm NS}))^{-1}\\
& = \ldots
% TODO: give expanded expression for this and move to earlier section
\end{eqnarray}
via the {\tt mdot} method, and incorporating the estimated distance distribution already determined. 
%
Based on the distance constraints, the estimated accretion rate  is within the range $(2.0_{-0.4}^{+0.6})\times10^3\ \rm{g\,cm^{-2}\,s^{-1}}$ for the usual assumed range for neutron star mass and radius (and hence redshift).

@@

Instead, we can estimate the ignition column using equation \ref{eq:column}, given an estimate of the nuclear burning yield, $\qnuc$. Since this quantity is unkown, we assume a set of randomly generated values of \Xb\
to estimate $\yign$ from equation \ref{eq:column}
%
We choose a uniform distribution for \Xb\ in the range 0--0.7; this distribution is probably not realistic for a wide sample of bursts, but it is difficult to do better.
% text copied from Kuiper et al. 2020 down to ********
% For the single burst observed from IGR~J17591$-$2342 
The inferred ignition column is $2.2_{-0.8}^{+1.4}\times10^8\,{\rm g\,cm^{-2}}$, and the average expected recurrence time is $1.8\pm0.7$~d. 
% Such long recurrence times likely guarantee that the hydrogen in the accreted fuel is exhausted at the base prior to ignition, and explaining the fast rise and short duration of the burst. 
, and $\Delta t_{\rm rec} = \yign/\dot{m}$.
% not yet edited (need to modify sufficiently to avoid plagiarism issues
Given each pair of (\Xb,  $\Delta t_{\rm rec}$) we can infer the accreted H-fraction $X_0$, based on 
% the expectation that hydrogen burns via the hot-CNO cycle and will be exhausted in a time $t_{\rm CNO} =9.8(X_0/0.7)(Z_{\rm CNO}/0.02)^{-1}$ \citep[e.g.][]{lampe16}.
equation \ref{eq:X_0}.
%
Since $Z_{\rm CNO}$ is also unknown, we similarly draw random values from a uniform distribution in the range 0.0--0.02.
We then select only those $X_0$ values within a physically realistic range $X_0<0.75$, and compute the confidence limits on the remaining parameters. 
%
We can also infer lower (upper) limits on $X_0$ ($Z_{\rm CNO}$), although these limits are not strongly constraining; we find $X_0>0.17$ and $Z_{\rm  CNO}<0.017$ at 95\% confidence.
% *******

>>>>

Thermonuclear (type-I) bursts are observed from neutron stars accreting from low-mass ($\lesssim 1\ M_\odot$) binary companions \cite[e.g.][]{gal21a}. ...

@@

\todo{insert table with key burst parameters here? }

@@

One of the most significant astrophysical uncertainties that has largely been neglected to date is the anisotropy of the burst (and persistent) emission.
% [{\bf move discussion from later section to here?}]
% \subsection{Emission anisotropy}
% \label{subsec:anisotropy}
\todo{ bit more description here }

@@

From data taken prior to the burst ignition we can measure the persistent flux $F_{\rm per}$, a measure of the accretion rate; \todo{ add expression here; see eq. \ref{eq:fper}}
% and provided more than one burst is observed, we may also measure (or estimate) the recurrence time, $\Delta t$. 
the persistent spectra of burst sources are  more complex than the bursts themselves, and typically exhibit thermal as well as non-thermal components [{\bf ref}].
%
These parameters may be combined to estimate the $\alpha$-value, the (observed) ratio of the burst to persistent flux: 
\begin{equation}
\alpha = \frac{\Delta t F_{\rm per} c_{\rm bol}}{E_B}
\end{equation}
where $c_{\rm bol}$ is the bolometric correction giving the inverse fraction of total persistent flux emitted in the instrumental band.
%
The $\alpha$ parameter is an important diagnostic of the burst fuel and ignition conditions.
%
\todo{mention the average method, as used in the rates paper }

@@

Of the model input parameters, we can also infer the persistent flux level expected given the model-assumed accretion rate $\dot{m}$, as follows: % gal03d, equation 2
\begin{equation}
% function fper in burstclass.py
F_{p,\infty} = \frac{L_p}{4\pi d^2} = 
% \frac{\dot{M} Q_{\rm grav}}{1+z} \left(\frac{R_{\rm NS}}{d}\right)^2 \xi_p^{-1}
% expression used in code
\frac{\dot{m} Q_{\rm grav}}{4\pi d^2 (1+z)\xi_p c_{\rm bol}}
\label{eq:fper}
\end{equation}
% Q_grav already defined in the previous section
% where $Q_{\rm grav} = c^2z/(1+z) \approx GM_{\rm NS}/R_{\rm NS}$ is the gravitational energy release per gram. % gal03d 
%
The bolometric correction $c_{\rm bol}$ accounts for the expermental limitation that the persistent flux can only be measured over a limited instrumental passband. The bolometric correction is the ratio of the estimated bolometric flux to the band-limited value.

@@

Describe how the code is written and how it works, the astropy.distribution thingy, MCMC for error determination etc.
\begin{itemize}
  \item Approach for calculations via astropy distributions
  \item 7(?) different methods for defining observational/input quantities
\end{itemize}

@@

Here we describe different approaches that have been used to deduce system parameters from observations of thermonuclear bursts.
%
We describe how the {\sc concord} code can be used to fully account for the astrophysical uncertainties, and as a companion to this section (and the code itself) we provide a companion {\sc jupyter} notebook [{\bf ref}] which replicates the analyses described below.

% simple example

As a simple example of the code utility, we describe here how the confidence range for peak luminosity may be estimated for a burst from a source with independently known distance. We consider the PRE burst observed from 4U~0513$-$40 with the {\it Rossi X-ray Timing Explorer} ({\it RXTE}) PCA on MJD~54043.68857 \cite[observation ID 92403-01-15-04; MINBAR ID \#3443;][]{minbar}. According to the MINBAR analysis this burst reached a peak flux of $(21.7\pm0.6)\times10^{-9}\ \epcs$. The source is located in the globular cluster NGC~1851, for which the distance is estimated at $(10.32^{+0.20}_{-0.24})$~kpc \cite[]{WatkinsEtAl2015}.

The isotropic luminosity can be calculated from the peak flux and distance using the {\tt luminosity} method, with {\tt isotropic=True} as $2.77\times10^{38}\ \epcs$. 
%
However, a more accurate calculation should take into account the uncertainty in the flux; the (asymmetric) uncertainty in the distance; and the likely effect of the anisotropy in the burst emission, based on the estimated system inclination $i>80^\circ$ \cite[]{fiocchi11}. This can be achieved by passing the flux and distance uncertainties to the routine; {\sc astropy} distributions are generated for both these parameters, with the required properties (e.g. Figure \ref{fig:burstlum}, panel (a)). We also pass the inclination limits, and an array of inclination values is generated assuming an isotropic distribution of inclinations within the limits. We calculate the anisotropy factor based on the modelling of \cite{he16}, and the user can select different models to quantify the variation introduced.
%
The resulting luminosity is significantly higher, $(4.85_{-0.42}^{+0.46})\times10^{38}\ \eps$ (Fig. \ref{fig:burstlum}, panel (b)); the discrepancy is primarily the result of the significant attenuation of the burst flux from the viewing angle close to the plane of the accretion disk. The corresponding anisotropy value $\xi_b$ is in the range 1.56--2 (Fig. \ref{fig:burstlum}, panel (c)). The dependence of the luminosity on the assumed inclination is illustrated in (Fig. \ref{fig:burstlum}, panel (d)).

The output from the function is by default the median value and the lower and upper uncertainties, at the required confidence level (68\% by default). Using the {\tt fulldist=True} flag, the function instead returns a dictionary including the full distributions for the luminosity, distance, inclination, anisotropy factor, as well as the anisotropy model identifier. % and the confidence level
Each of the distribution arrays can be accessed via the {\tt distribution} attribute.
This object can be written to a file or used for input to subsequent analyses.

\begin{figure}[ht!]
\plotone{fig0.pdf}
\caption{Illustration of calculation of burst peak luminosity using {\sc concord}, incorporating measurement uncertainties and the modeled effect of the burst anisotropy. The adopted distance PDF is shown in panel (a); note the slight skew in the distribution. The resulting luminosity PDF in panel (b) is substantially higher than the isotropic value ({\it red dashed line}). The $1\sigma$ confidence intervals are indicated by the green lines. The PDF for the anisotropy factor $\xi_b$ is shown in panel (c); the dependence of the luminosity on the adopted system inclination is shown in panel (d).
\label{fig:burstlum}}
\end{figure}

[{\bf comment here perhaps on the additional uncertainties that might contribute; e.g. for $M$, $R$; code is flexible enough to incorporate them }]

In the sections below we describe other examples,
%
focussing on adopting approaches tailored to the availability of data, which will differ for different sources.

@@

The distance can be estimated from the peak flux of a PRE burst based on equation \ref{disteq}, using the {\tt dist} method. The challenge is taking into account the uncertainties in the various parameters that go into this calculation. If we adopt the empirical Eddington luminosity of \cite[]{kuul03a}, the only remaining factor is the burst emission anisotropy $\xi_b$.
%
As for the case of the burst in IGR~J00291+5934, the approach adopted in {\sc concord} to take this into account is to generate a distribution of $\xi_b$ values, based on an isotropic distribution of inclinations within known limits. The presence (absence) of X-ray dips occurring at the binary orbital period will constrain the inclination to greater (less) than $75^\circ$ [{\bf ref}]. However, more stringent constraints may be available.

@@

Following \cite{kuiper20}, we estimate the distance to the source based on the empirical value of the Eddington luminosity, and incorporating the expected anisotropy for an inclination range of 
% kuiper20 uses slightly more restrictive 28-30
24--30$^\circ$ \cite[]{sanna18} as $7.7_{-0.6}^{+0.8}$~kpc. Here the uncertainty is dominated by the peak flux, since the anisotropy factor is effectively fixed for such a narrow range of inclination, at $\xi_b=0.70$.
%
The PDF for the distance, which is distinctly non-Gaussian, is plotted in Figure \ref{fig:dist_igr17591}. We also show the corresponding distribution resulting from adoption of the theoretical Eddington luminosity, according to equation \ref{ledd}.

@@

The {\sc concord} code allows us to pass the derived distance distribution directly to the method for calculating the accretion rate at the time of the burst, also incorporating the dependence of the persistent flux anisotropy (which is not the same as for the burst flux).
%
The persistent flux can then be converted to an accretion rate, given the estimated anisotropy factor predicted (separately) for the persistent emission, as
\begin{eqnarray}
\dot{m} & = L_{\rm pers} (1+z) (4\pi R^2(GM/R_{\rm NS}))^{-1}\\
& = \ldots
% TODO: give expanded expression for this and move to earlier section
\end{eqnarray}
via the {\tt mdot} method, and incorporating the estimated distance distribution already determined. 
%
Based on the distance constraints, the estimated accretion rate  is within the range $(2.0_{-0.4}^{+0.6})\times10^3\ \rm{g\,cm^{-2}\,s^{-1}}$ for the usual assumed values for neutron star mass and radius (and hence redshift).

@@

Instead, we can constrain the fuel composition via the ignition column (equation \ref{eq:column}), based on an estimate of the nuclear burning yield, $\qnuc$. Since this quantity is unknown, we assume a set of random uniformly-distributed\footnote{Such distribution may not be realistic, but it serves to demonstrate the code capabilities. The largest (and most unlikely) values of \Xb\ are suppressed by the limit on $X_0$ imposed during this process.} values of \Xb\ in the range 0--0.7
to estimate $\yign$ from equation \ref{eq:column}.
%
We then estimate the recurrence time for each pair of $\yign$, $\dot{m}$ values as
\begin{equation}
\Delta t_{\rm rec} = (1+z)\yign/\dot{m}
\end{equation}
And finally calculate the fuel composition, $X_0$, for each set of \Xb, $\yign$, $\Delta t$ from equation \ref{eq:X_0}, again assuming a uniform distribution for $Z_{\rm CNO}$ between 0--0.02.
%
Because we have chosen uninformative prior distributions for \Xb\ and $Z_{\rm CNO}$, some of the $X_0$ values are unrealistic, well in excess of the maximum expected for such sources, 0.75 or so. We thus limit the distributions of each of the parameters to those such that $X_0\leq0.75$, and quote the $1\sigma$ parameter ranges for each.

% We choose a uniform distribution for \Xb\ in the range 0--0.7; this distribution is % probably not realistic for a wide sample of bursts, but it is difficult to do better.
% text copied from Kuiper et al. 2020 down to ********
% For the single burst observed from IGR~J17591$-$2342 
% Original values; couldn't replicate in 2021 Sep
% The inferred ignition column is $2.2_{-0.8}^{+1.4}\times10^8\,{\rm g\,cm^{-2}}$, and the average expected recurrence time is $1.8\pm0.7$~d. 
% 2021 Sep values
The inferred ignition column is $1.63_{-0.58}^{+1.05}\times10^8\,{\rm g\,cm^{-2}}$, and the average expected recurrence time is $1.19_{-0.42}^{+0.79}$~d. 
% Such long recurrence times likely guarantee that the hydrogen in the accreted fuel is exhausted at the base prior to ignition, and explaining the fast rise and short duration of the burst. 
% not yet edited (need to modify sufficiently to avoid plagiarism issues
%
We can also infer lower (upper) limits on $X_0$ ($Z_{\rm CNO}$), although these limits are not strongly constraining; we find $X_0>0.17$ and $Z_{\rm  CNO}<0.017$ at 95\% confidence.
% *******
